Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
Abstract: fix typo.
Section 1: Add a TODO about travel award.
Updated NCH's keynote description

\begin{abstract}      
This technical report records and discusses the Second Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE2). The workshop used an
alternative submission and peer-review process, which led to a set of papers
divided across five topic areas: exploring sustainability; software development
experiences; credit \& incentives; reproducibility \& reuse \& sharing; and code
testing \& code review. The report includes a description of the submission and
review process, the workshop keynote presentations, a series of lightning talks, a discussion on
sustainability, and the five discussions from the topic areas. The report also
contains a description of potential actions that were proposed in the the five
discussions.
\end{abstract}

@@

Neil Chue Hong is currently serving as the Director of the Software
Sustainability Institute (SSI), United Kingdom. Chue Hong started with making
the point that world-leading research relies on software, quoting Kersten Kleese
Van Dam of the Pacific Northwestern National Laboratory via a petition campaign at \katznote{add cite for the petition} 
change.org, ``\emph{Today there are very few science
areas left which do not rely on IT and thus software for the majority of the
their research work. More importantly key scientific advances in experimental
and observational science would have been impossible without better software.}''
He also cited Daniel Katz, Software Infrastructure for Sustained Innovation
(SI2) Program Director of the National Science Foundation, ``\emph{Scientific
discovery and innovation are advancing along fundamentally new pathways opened
by development of increasingly sophisticated software. Software is an integral
enabler of computation, experiment and theory, and directly responsible for
increased scientific productivity and enhancement of researchers'
capabilities.}''

Chue Hong drew attention to the fact that we too often disregard hundreds of
thousands of software researchers thinking they are the long tail. We forget
that actually they are the mainstream and we are the elite minority. He
emphasized that software is no longer special; it is both essential to and
common in scientific research. A survey of researchers from 15 Russell Group
universities conducted by the SSI between Aug--Oct 2014, which included a total
of 406 respondents covering a representative range of funders, disciplines, and
seniority, reported that 92\% confirmed the use of research software such as
MATLAB, SPSS, Excel, etc. and 89\% affirmed that it would be impossible or
difficult to conduct research without software. Nevertheless the British
research community is just starting to understand the magnitude of the issue. In
the aforementioned SSI survey, 56\% of researchers developed their own research
software, 71\% had no formal software-development training, 140,000 researchers
relied on their own coding skills, and only 4\% of jobs advertised in British
universities were software related.

According to Chue Hong, costs of software-reliant research in the United Kingdom
included \textsterling 840 million of investment in the financial year
2013--2014, and this amount has risen by 3\% on average over the past four
years. About 30\% of total research investment has been spent on research that
relies on software over the last four financial years. These numbers stemmed
from an analysis of data from $49,650$ grant titles and abstracts published on
Gateway to Research between years 2010 and 2014.

Chue Hong led the audience in discussing the following questions: What are we
going to \emph{do} to help and benefit software researchers and to achieve
software sustainability if~92\% of researchers and WSSSPE care about scientific
software? Who do we need to persuade? What are the incentives we need to put in
place? He challenged the workshop participants to change the current deficient
practices in research and academia.

\begin{comment}
With apologies to Greg Wilson:
\href{http://software-carpentry.org/blog/2013/10/you-keep-using-that-word.html}
{You Keep Using That Word}
We can talk all day about the best ways of achieving sustainability
But what are we going to do to make it actually sustainable?

Careers outside academic sector: 
Non-university Research (industry, government, etc.)
UK STEM graduate career paths: Early Career Research, Permanent Research Staff,
Professor.
The Scientific Century, Royal Society, 2010 (revised to reflect first stage
clarification from ``What Do PhD's Do?'' study.)
Who	are the people  we need to persuade?
What	 are	the incentives we need to put in place?
How are we going to show that the work we present here has benefit to the 92\%?
We are all the 92\%	and we can change the world if we	want!
\end{comment}

>>>>

\begin{abstract}      
This technical report records and discusses the Second Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE2). The workshop used an
alternative submission and peer-review process, which led to a set of papers
divided across five topic areas: exploring sustainability; software development
experiences; credit \& incentives; reproducibility \& reuse \& sharing; and code
testing \& code review. The report includes a description of the submission and
review process, the workshop keynote presentations, a series of lightning talks, a discussion on
sustainability, and the five discussions from the topic areas. The report also
contains a description of potential actions that were proposed in the five
discussions.
\end{abstract}

@@

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom. The SSI was founded to support the UK?s research software 
community by cultivating better, more sustainable, research software to enable 
world-class research. Chue Hong's keynote\footnote{Slides available at: http://dx.doi.org/10.6084/m9.figshare.1243288}
 started by making the point of that the use of 
- and reliance on - software is pervasive in all areas of world-leading research, 
showing examples from disciplines as diverse as the humanities and high-energy 
physics, and quoting Kersten Kleese Van Dam of the Pacific Northwestern 
National Laboratory via a petition campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}} 
at change.org, ``\emph{Today there are very few science areas left 
which do not rely on IT and thus software for the majority of their research work. 
More importantly key scientific advances in experimental and observational science 
would have been impossible without better software.}'' He also cited Daniel Katz, 
Software Infrastructure for Sustained Innovation(SI2) Program Director of the 
National Science Foundation, ``\emph{Scientific discovery and innovation are 
advancing along fundamentally new pathways opened by development of 
increasingly sophisticated software. Software is an integral enabler of 
computation, experiment and theory, and directly responsible for increased 
scientific productivity and enhancement of researchers' capabilities.}''

Chue Hong drew attention to the issue that in the cyber-infrastructure and high 
performance community, we all too often disregard the hundreds of thousands 
of researchers developing software thinking they are the long tail. In fact, the 
numbers point to the fact that they are actually the mainstream and we (referring 
to people including the audience of the talk) are the elite minority. He
emphasized that software is no longer special; it is both essential to and
common in scientific research. A 2014 survey\footnote{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers} 
conducted by the SSI polled researchers 
from 15 research-intensive UK universities (406 respondents covering a 
representative range of funders, disciplines, and seniority) reported that 92\% 
confirmed the use of research software  and 89\% affirmed that it would be 
impossible or difficult to conduct research without software. Nevertheless the British
research community is just starting to understand the magnitude of the issue. 
Whilst many researchers make use of software such as MATLAB, SPSS, and 
Excel, data from the aforementioned SSI survey shows that over half (56\%) of 
respondents developed their own research software (which equates to over 
140,000 researchers if extrapolated across the UK) and yet 71\% had no formal 
software-development training, having to rely on their own coding skills.

Examining another aspect of the size of the research software community, 
Chue Hong noted that the costs of software-reliant research in the United Kingdom
included \textsterling 840 million of investment in the financial year
2013--2014, and this amount has risen by 3\% on average over the past four
years. About 30\% of total research investment has been spent on research that
relies on software over the last four financial years. These numbers stemmed
from an analysis by the SSI of data from $49,650$ grant titles and abstracts published on
Gateway to Research between years 2010 and 2014. A similar analysis of university 
jobs advertised in the same period discovered that despite this investment, only 4\% of 
positions were software development related, and of these only 17\% were explicitly 
named as a software developer / software engineer position: the vast majority being 
advertised as research associate / research assistant positions. This in turn leads to
the issue of career paths for those bridging the research and software worlds, who
are essential to support the use and further development of research software, a point 
highlighted by a graphic showing UK STEM graduate career paths\footnote{Source: The Scientific Century, Royal Society, 2010 (revised to reflect first stage clarification from ``What Do PhD's Do?'' study)}
 showing that only 3.5\% were able to secure permanent positions.

To conclude, Chue Hong led the audience in discussing the following questions: 
What are \emph{we} going to \emph{do} to help and benefit the 92\% of researchers 
who rely on software? Who do we need to persuade? What are the incentives we 
need to put in place? Finally, he challenged the workshop participants to change 
the current deficient practices in research and academia.
